---
title: "rsa_april"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
V_vals <- seq(0.0001, 0.9999, length.out = 2000) 

#grid of beta params
mu_vals <- seq(0.06, 0.94, length.out = 12)
nu_vals <- seq(1, 10, length.out = 10)

sample_mu_vals <- c(0.25, 0.5, 0.75)

#might make more conceptual sense to restrict these to higher values 
#but I think it belongs in the L1 prior rather enforcing it here
theta_vals <- seq(0.05, 0.95, length.out = 46)

#chosen pretty arbitrarily, needs further optimization
cost <- c(null = 0, long = 0.2) 
alpha <- 2
```

Literal listener:

$$
P_{L_0^{\mu, \nu}}(V|R) \propto P(R|V)\cdot P(V) \\
=  P(R|V)\cdot Beta(V; \mu \nu, (1-\mu)\nu)
$$

```{r}
#convert from mean/sample size parameterization 
#back to alpha/beta (so i can use dbeta)
#again note that mu/nu are not inferred by L0
#that's just the distribution that S1 is treating as his prior
#L1 is inferring /which beta distribution/ S1 is treating as L0's prior 
prior_L0 <- function(mu, nu) {
  a <- mu*nu
  b <- (1-mu) * nu
  dbeta(V_vals, a, b)
}

#i picked the linear likelihood for simplicity
#also tried quadratic to make the reasons have a bigger effect
#i probably need to defend making the likelihood independent of the 
#hyperparameters of the prior? 
likelihood_L0 <- function(R) {
  if(R == "null") rep(1, length(V_vals)) else V_vals^2
}

lit_listener <- function(mu, nu, R) {
  prior <- prior_L0(mu, nu)
  prior <- prior / sum(prior)
  likelihood <- likelihood_L0(R)
  unnorm_posterior <- prior * likelihood
  posterior <- unnorm_posterior / sum(unnorm_posterior)
  data.frame(V = V_vals, 
             Prior = prior, 
             Posterior = posterior,
             UnnormPosterior = unnorm_posterior,
             mu = mu, 
             nu = nu, 
             R = R)
}
```

```{r}

L0_data <- lapply(mu_vals, function(m) 
            lapply(nu_vals, function(n) lit_listener(m, n, "long"))) %>%
            bind_rows() %>%
            pivot_longer(cols = c("Prior", "Posterior"), names_to = "Distribution",values_to ="Prob")

figure1_data <- L0_data  %>%
  filter(Distribution == "Posterior") %>%
  group_by(mu, nu) %>%
  summarise(post_mean = sum(V * Prob), .groups = "drop") %>%
  mutate(delta_means = post_mean - mu)

```

```{r}

figure1_data %>% ggplot(aes(x = mu, y = nu, fill = delta_means)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(title = "Change in expected value from null (0R) to long (4R) explanation", x = "Prior mu", y = "Prior nu", fill = "Delta") +
  theme_minimal()

```

The general trend we observe here is that there's always a positive change from the prior to the posterior, and that tends to be magnified when the prior is less favorable on average (low mu) or less confident (low nu). Since the null utterance reproduces the prior, and the long utterance conditions on the evidence, the long utterance always does better than the null utterance for L0.

For good measure, for fixed nu:

```{r}

figure2_data <- lapply(sample_mu_vals, function(m) lit_listener(m, 6, "long")) %>%
            bind_rows() %>%
            pivot_longer(cols = c("Prior", "Posterior"), names_to = "Distribution",values_to ="Prob")

posterior_means <- figure2_data %>%
  filter(Distribution == "Posterior") %>%
  group_by(mu) %>%
  summarise(post_mean = sum(V * Prob))

figure2_plot <- ggplot(figure2_data, aes(x = V, y = Prob, fill = Distribution, color = Distribution)) +
  geom_area(alpha = 0.3, position = "identity") + 
  geom_line(size = 0.8) +
  geom_vline(aes(xintercept = mu), linetype = "dashed", color = "gray30", linewidth = 0.5)+ 
  geom_vline(data = posterior_means, aes(xintercept = post_mean), linetype = "dashed", color = "blue", linewidth = 0.5)+
  facet_wrap(~mu, labeller = label_both, ncol=1) +
  scale_fill_manual(values = c("Prior" = "gray50", "Posterior" = "blue")) +
  scale_color_manual(values = c("Prior" = "gray30", "Posterior" = "blue")) +
  theme_bw() +
  labs(title = "Bayesian update with no pragmatic reasoning", x = "Value", y = "P(V|long-4R; mu, nu)")

print(figure2_plot)
```

Here also we can see that the distribution always shifts to the right for the long explanation compared to the prior (no explanation), but the shift is less when the prior is already high.

```{r}


```

```{r}
results_null <- mapply(function(m, n) lit_listener(m, n, "null")$Posterior,
  rep(mu_vals, times = length(nu_vals)),
  rep(nu_vals, each = length(mu_vals))
)
results_long <- mapply(function(m, n) lit_listener(m, n, "long")$Posterior,
  rep(mu_vals, times = length(nu_vals)),
  rep(nu_vals, each = length(mu_vals))
)
results_null_unnorm <- mapply(function(m, n) lit_listener(m, n, "null")$UnnormPosterior,
  rep(mu_vals, times = length(nu_vals)),
  rep(nu_vals, each = length(mu_vals))
)
results_long_unnorm <- mapply(function(m, n) lit_listener(m, n, "long")$UnnormPosterior,
  rep(mu_vals, times = length(nu_vals)),
  rep(nu_vals, each = length(mu_vals))
)

L0_array_null <- array(NA, dim = c(length(V_vals), length(mu_vals), length(nu_vals)))
L0_array_long <- array(NA, dim = c(length(V_vals), length(mu_vals), length(nu_vals)))
L0_array_null_unnorm <- array(NA, dim = c(length(V_vals), length(mu_vals), length(nu_vals)))
L0_array_long_unnorm <- array(NA, dim = c(length(V_vals), length(mu_vals), length(nu_vals)))

for (k in seq_along(nu_vals)) {
  for (j in seq_along(mu_vals)) {
    idx <- (k - 1) * length(mu_vals) + j
    L0_array_null[, j, k] <- results_null[, idx]
    L0_array_long[, j, k] <- results_long[, idx]
    L0_array_null_unnorm[, j, k] <- results_null_unnorm[, idx]
    L0_array_long_unnorm[, j, k] <- results_long_unnorm[, idx]
  }
}

speaker_grid <- expand.grid(
  theta_i = 1:length(theta_vals), 
  mu_j = 1:length(mu_vals), 
  nu_k = 1:length(nu_vals)
)

speaker_grid$theta <- theta_vals[speaker_grid$theta_i]
speaker_grid$mu <- mu_vals[speaker_grid$mu_j]
speaker_grid$nu <- nu_vals[speaker_grid$nu_k]
```

Pragmatic speaker

$$
P_{L_0^{\mu,\nu}}(O > \theta| R) = \int_\theta^1 P_{L_0^{\mu, \nu}}(O|R)  dO
$$

$$
\mathbb{U}_{S_1}(R; \theta, \mu, \nu) = \underbrace{\log P_{L_0^{\mu, \nu}}(O > \theta|R)}_{\textbf{-surprisal}} - \text{cost(R)}
$$

$$
P_{S_1}(R|\theta, \mu) = \text{softmax}(\alpha \cdot \mathbb{U}_{S_1}(R; \theta, \mu))
$$

```{r}

s_probs <- mapply(function(i, j, k) {
  
  #integral
  p_null <- sum(L0_array_null[V_vals > theta_vals[i], j, k])
  p_long <- sum(L0_array_long[V_vals > theta_vals[i], j, k])
  
  #informativity 
  inf_null <- log(p_null + 1e-10)
  inf_long <- log(p_long + 1e-10)
  
  #utility
  util_null <- inf_null - cost["null"]
  util_long <- inf_long - cost["long"]

  #softmax
  null_numer <- exp(alpha * util_null)
  long_numer <- exp(alpha * util_long)
  denom <- null_numer + long_numer
  
  return(c(null_numer/denom, long_numer/denom, p_null, p_long, inf_null, inf_long))
  
}, speaker_grid$theta_i, speaker_grid$mu_j, speaker_grid$nu_k)

speaker_grid$prob_null <- s_probs[1, ]
speaker_grid$prob_long <- s_probs[2, ]
speaker_grid$int_null <- s_probs[3, ]
speaker_grid$int_long <- s_probs[4, ]
speaker_grid$inf_null <- s_probs[5, ]
speaker_grid$inf_long <- s_probs[6, ]


speaker_grid <- speaker_grid %>%
  mutate(int_diff = int_long - int_null) %>%
  mutate(inf_diff = inf_long - inf_null) %>%
  mutate(prob_diff = prob_long - prob_null) %>%
  filter(theta_i == 10)
```

```{r}

ggplot(speaker_grid, aes(x = mu, y = nu, fill = int_diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(x = "mu", y = "nu", fill = "diff") +
  theme_minimal()


ggplot(speaker_grid, aes(x = mu, y = nu, fill = prob_diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(x = "mu", y = "nu", fill = "diff") +
  theme_minimal()

```

These results are generally as expected. the long utterance is preferred over the null for lower prior mean, and the null utterance is preferred for higher prior mean.

Further, the long utterance survives for higher values of mu when nu is lower (weakly opinionated prior).

```{r}

prior_theta <- dnorm(theta_vals, mean = 0.5, sd = 0.2)
prior_theta <- prior_theta / sum(prior_theta)

prior_mu_uniform <- rep(1, length(mu_vals)) / length(mu_vals)
prior_nu_uniform <- rep(1, length(nu_vals)) / length(nu_vals)

mu_probs <- prior_mu_uniform[speaker_grid$mu_j]
nu_probs <- prior_nu_uniform[speaker_grid$nu_k]
theta_probs <- prior_theta[speaker_grid$theta_i]

joint_null <- speaker_grid$prob_null * theta_probs * mu_probs * nu_probs
joint_long <- speaker_grid$prob_long * theta_probs * mu_probs * nu_probs

group_key <- speaker_grid$mu_j * 1000 + speaker_grid$nu_k

marg_null <- rowsum(joint_null, group_key)
marg_long <- rowsum(joint_long, group_key)
marg_null <- marg_null / sum(marg_null)
marg_long <- marg_long / sum(marg_long)

mu_nu_grid <- speaker_grid %>%
  distinct(mu_j, nu_k, mu, nu) %>%
  arrange(mu_j, nu_k) %>%
  mutate(group_key = mu_j * 1000 + nu_k)

null_prob <- sapply(seq_len(nrow(mu_nu_grid)), function(idx) {
  L0_array_null[, mu_nu_grid$mu_j[idx], mu_nu_grid$nu_k[idx]]
}) %*% marg_null
null_prob <- null_prob / sum(null_prob)

long_prob <- sapply(seq_len(nrow(mu_nu_grid)), function(idx) {
  L0_array_long[, mu_nu_grid$mu_j[idx], mu_nu_grid$nu_k[idx]]
}) %*% marg_long
long_prob <- long_prob / sum(long_prob)

L1_marginals <- data.frame(
  mu = mu_nu_grid$mu,
  nu = mu_nu_grid$nu,
  marg_null = as.numeric(marg_null),
  marg_long = as.numeric(marg_long)
)

L1_posterior <- rbind(
  data.frame(V = V_vals, Prob = as.numeric(null_prob), Distribution = "null"),
  data.frame(V = V_vals, Prob = as.numeric(long_prob), Distribution = "long")
)
```

```{r}

L1_marginals_plot <- L1_marginals %>%
  mutate(marg_diff = marg_long - marg_null) 

ggplot(L1_marginals_plot, aes(x = mu, y = nu, fill = marg_diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(x = "mu", y = "nu", fill = "diff") +
  theme_minimal()

```

This is also generally as expected. we see more mass on the long utterance vs the null (diff is high/blue) for lower values of mu, and less mass on the long utterance vs

the null (diff is low/red) for higher values of mu. again, long survives for higher values of mu when nu is lower (weakly opinionated prior)

```{r}

prior_theta <- dnorm(theta_vals, mean = 0.5, sd = 0.2)
prior_theta <- prior_theta / sum(prior_theta)

nu_scaled <- (nu_vals - min(nu_vals)) / (max(nu_vals) - min(nu_vals))
nu_scaled <- pmax(pmin(nu_scaled, 0.999), 0.001)

group_key <- speaker_grid$mu_j * 1000 + speaker_grid$nu_k
theta_probs <- prior_theta[speaker_grid$theta_i]

mu_nu_grid <- speaker_grid %>%
  distinct(mu_j, nu_k, mu, nu) %>%
  arrange(mu_j, nu_k) %>%
  mutate(group_key = mu_j * 1000 + nu_k)

L0_null_mat <- sapply(seq_len(nrow(mu_nu_grid)), function(idx) {
  L0_array_null[, mu_nu_grid$mu_j[idx], mu_nu_grid$nu_k[idx]]
})
L0_long_mat <- sapply(seq_len(nrow(mu_nu_grid)), function(idx) {
  L0_array_long[, mu_nu_grid$mu_j[idx], mu_nu_grid$nu_k[idx]]
})

L1_function <- function(prior_mu, prior_nu) {
  
  current_p_mu <- dbeta(mu_vals, prior_mu * 3, (1 - prior_mu) * 3)
  current_p_mu <- current_p_mu / sum(current_p_mu)
  
  prior_nu_scaled <- (prior_nu - min(nu_vals)) / (max(nu_vals) - min(nu_vals))
  prior_nu_scaled <- pmax(pmin(prior_nu_scaled, 0.999), 0.001)
  current_p_nu <- dbeta(nu_scaled, prior_nu_scaled * 3, (1 - prior_nu_scaled) * 3)
  current_p_nu <- current_p_nu / sum(current_p_nu)
  
  mu_probs <- current_p_mu[speaker_grid$mu_j]
  nu_probs <- current_p_nu[speaker_grid$nu_k]
  
   # 1. compute joints — these retain relative scaling between null and long
  joint_null <- speaker_grid$prob_null * theta_probs * mu_probs * nu_probs
  joint_long <- speaker_grid$prob_long * theta_probs * mu_probs * nu_probs
  
  # marg_null and marg_long: normalize separately (they are P(mu,nu|R) for each R)
  marg_null <- rowsum(joint_null, group_key)
  marg_long <- rowsum(joint_long, group_key)
  marg_null <- marg_null / sum(marg_null)
  marg_long <- marg_long / sum(marg_long)
  
  marg_null <- marg_null[as.character(mu_nu_grid$group_key), , drop = FALSE]
  marg_long <- marg_long[as.character(mu_nu_grid$group_key), , drop = FALSE]
  
  # use unnormalized L0 as P(R|V) * P(V; mu, nu)
  null_prob <- as.numeric(L0_null_mat %*% marg_null)
  long_prob <- as.numeric(L0_long_mat %*% marg_long)
  
  # normalize each separately at the end — the proportionality in the equation
  null_prob <- null_prob / sum(null_prob)
  long_prob <- long_prob / sum(long_prob)
  
  ev_null <- sum(V_vals * null_prob)
  ev_long <- sum(V_vals * long_prob)
  
  data.frame(prior_mu = prior_mu, prior_nu = prior_nu, ev_diff = ev_long - ev_null)
}

L1_results <- do.call(rbind, lapply(mu_vals, function(pm)
  do.call(rbind, lapply(nu_vals, function(pn)
    L1_function(pm, pn)))))
```

```{r}
ggplot(L1_results, aes(x = prior_mu, y = prior_nu, fill = ev_diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(x = "prior mu", y = "prior nu", fill = "E[V|long] - E[V|null]") +
  theme_minimal()
```

```{r}
L1_results %>%
  filter(prior_nu == nu_vals[10]) %>%
  arrange(prior_mu) %>%
  select(prior_mu, ev_diff)
```

```{r}

sum(L0_null_mat)
sum(L0_long_mat)
```

```{r}

```
