---
title: "rsa_march"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
O_vals <- seq(0.01, 0.99, length.out = 50) 
mu_vals <- seq(0.05, 0.95, length.out = 50)
theta_vals <- seq(0.5, 0.9, length.out = 50) #i feel like this range conceptually makes sense but might be creating issues with the model
sample_mu_vals <- c(0.2, 0.5, 0.8)

#ive been fussing with these but i think the problem might just be the model design itself and not the tuning
#and also just doing a hyperparameter sweep to find what gives the effect i want feels a bit fraudulent lol
cost <- c(null = 0, long = 0.2) 
alpha <- 10
nu <- 6 #keeping this constant here for simplicity but a better version should infer both parameters
```

```{r}
#convert from mean/sample size back to alpha/beta
#again note that mu/nu are not inferred by L0, that is just his prior
#L1 is inferring /which beta distribution/ S1 is treating as L0's prior 
prior_L0 <- function(mu, nu) {
  a <- mu*nu
  b <- (1-mu) * nu
  dbeta(O_vals, a, b)
}

#linear likelihood for the long explanation case might be too weak?
#but i tried it with a quadratic too and that didn't fix it
likelihood_L0 <- function(R) {
  if(R == "null") rep(1, length(O_vals)) else O_vals
}

```

$$
P_{L_0}(O|R; \mu) \propto P(R|O)\cdot P(O;\mu) 
$$

```{r}
#calculating the above equation
lit_listener <- function(mu, nu, R) {
  prior <- prior_L0(mu,nu) 
  prior <- prior / sum(prior)
  likelihood <- likelihood_L0(R)
  posterior <- (prior * likelihood) / sum(prior * likelihood)
  data.frame(O = O_vals, Prior = prior, Posterior = posterior, mu = mu, R = R)
}
```

```{r}
figure1_data <- lapply(sample_mu_vals, function(m) lit_listener(m, nu, "long")) %>%
  bind_rows() %>%
  pivot_longer(cols = c("Prior", "Posterior"), names_to = "Distribution", values_to = "Prob")

posterior_means <- figure1_data %>%
  filter(Distribution == "Posterior") %>%
  group_by(mu) %>%
  summarise(post_mean = sum(O * Prob))

figure1_plot <- ggplot(figure1_data, aes(x = O, y = Prob, fill = Distribution, color = Distribution)) +
  geom_area(alpha = 0.3, position = "identity") + 
  geom_line(size = 0.8) +
  #i added the vline since the mean is not actually at the peak in a beta lol which i sort of forgot
  #and was maybe misinterpreting a bit in the achimova paper :)
  #but the intuition does seem to come out that there's less change at higher priors
  geom_vline(aes(xintercept = mu), linetype = "dashed", color = "gray30", linewidth = 0.5)+ 
  geom_vline(data = posterior_means, aes(xintercept = post_mean), linetype = "dashed", color = "blue", linewidth = 0.5)+
  facet_wrap(~mu, labeller = label_both, ncol=1) +
  scale_fill_manual(values = c("Prior" = "gray50", "Posterior" = "blue")) +
  scale_color_manual(values = c("Prior" = "gray30", "Posterior" = "blue")) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)))+
  labs(title = "Bayesian update with no pragmatic reasoning", x = "Opinion (O)", y = "P(O|R; mu)")

print(figure1_plot)

```



```{r}
#cached since the lookups were taking one century to run
L0_matrix_null <- sapply(mu_vals, function(mu) lit_listener(mu, nu, "null")$Posterior)
L0_matrix_long <- sapply(mu_vals, function(mu) lit_listener(mu, nu, "long")$Posterior)

speaker_grid <- expand.grid(
  theta_i = 1:length(theta_vals), 
  mu_j = 1:length(mu_vals)
)

speaker_grid$theta <- theta_vals[speaker_grid$theta_i]
speaker_grid$mu <- mu_vals[speaker_grid$mu_j]

dim(speaker_grid)
```

$$
P_{L_0}(O > \theta| R; \mu) = \int_\theta^1 P_{L_0}(O|R; \mu)  dO
$$

$$
\mathbb{U}_{S_1}(R; \theta, \mu) = \underbrace{\log P_{L_0}(O > \theta|R; \mu)}_{\textbf{-surprisal}} - \text{cost(R)}
$$ 

$$
P_{S_1}(R|\theta, \mu) = \text{softmax}(\alpha \cdot \mathbb{U}_{S_1}(R; \theta, \mu))
$$

```{r}

s_probs <- mapply(function(i, j) {
  
  #integral
  p_null <- sum(L0_matrix_null[O_vals > theta_vals[i], j])
  p_long <- sum(L0_matrix_long[O_vals > theta_vals[i], j])
  
  #utility
  util_null <- log(p_null + 1e-10) - cost["null"]
  util_long <- log(p_long + 1e-10) - cost["long"]

  #softmax
  null_numer <- exp(alpha * util_null)
  long_numer <- exp(alpha * util_long)
  denom <- null_numer + long_numer
  
  return(c(null_numer/denom, long_numer/denom))
  
}, speaker_grid$theta_i, speaker_grid$mu_j)

speaker_grid$prob_null <- s_probs[1, ]
speaker_grid$prob_long <- s_probs[2, ]

print(tail(speaker_grid, 100))
```

```{r}
target_theta <- 0.3
closest_theta <- theta_vals[which.min(abs(theta_vals - target_theta))] #just so i dont have to keep fiddling with it when im testing diff ranges of thetas

speaker_plot <- speaker_grid %>% 
  filter(theta == closest_theta) %>%
  pivot_longer(cols = c(prob_null, prob_long), names_to = "Utterance",values_to = "Probability") %>%
  mutate(Utterance = recode(Utterance, prob_null = "null", prob_long = "long"))%>% 
  mutate(Utterance = factor(Utterance, levels = c("null", "long")))

print(speaker_plot)
```

```{r}
ggplot(speaker_plot, aes(x = mu, y = Probability, color = Utterance)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = c("null" = "gray30", "long" = "blue")) +
  theme_classic() +
  labs(
    title = paste("Pragmatic speaker utterance probabilities for theta =", target_theta),
    x = "Possible priors (mu)",
    y = "P(R | theta, mu)",
    color = "Explanation"
  ) +
  ylim(0, 1)
```
$$
P_{L_1}(\theta, \mu|R) \propto P_{S_1}(R|\theta, \mu) \cdot P(\theta) \cdot P(\mu)
$$
$$
P_{L_1}(\mu|R) = \int_0^1  P_{L_1}(\theta, \mu|R) d\theta
$$
$$
P_{L_1}(O|R) \propto \int_0^1\underbrace{ P(R|O) \cdot P(O;\mu)}_{\textbf{Conditionalization from prior}}  \cdot \underbrace{P_{L_1}(\mu|R)}_{\textbf{Probability of that prior}} d\mu
$$

```{r}
# i have a sense that P(mu) and P(theta) are creating the problem here, just flagging that
prior_theta <- dnorm(theta_vals, mean = 0.7, sd = 0.2) #also tried this with a uniform prior
prior_theta <- prior_theta / sum(prior_theta)

L1_results <- lapply(sample_mu_vals, function(prior) {
  current_p_mu <- dbeta(mu_vals, prior*2, (1-prior)*2) #likewise tried a bunch of diff hyperpriors including uniform
  #so i can't really tell if it's just a hyperparameter tuning issue or something is fundamentally wrong with the logic of L1
  current_p_mu <- current_p_mu / sum(current_p_mu)

  mu_probs <- current_p_mu[speaker_grid$mu_j]
  theta_probs <- prior_theta[speaker_grid$theta_i]
  
  joint_null <- speaker_grid$prob_null * theta_probs * mu_probs
  joint_long <- speaker_grid$prob_long * theta_probs * mu_probs
  
  marg_null <- as.numeric(rowsum(joint_null, speaker_grid$mu))
  marg_long <- as.numeric(rowsum(joint_long, speaker_grid$mu))

  null_prob <- as.numeric(L0_matrix_null %*% marg_null)
  null_prob <- null_prob / sum(null_prob)
  
  long_prob <- as.numeric(L0_matrix_long %*% marg_long)
  long_prob <- long_prob / sum(long_prob)
  
  list(
    posterior = rbind(
      data.frame(O = O_vals, Prob = null_prob, mu = prior, Distribution = "null"),
      data.frame(O = O_vals, Prob = long_prob, mu = prior, Distribution = "long")
    ),
    marginals = data.frame(mu_val = mu_vals, marg_null = marg_null, marg_long = marg_long, prior = prior)
  )
}) 

posterior <- lapply(L1_results, function(x) x[["posterior"]]) %>% bind_rows()
marginals <- lapply(L1_results, function(x) x[["marginals"]]) %>% bind_rows()
```

```{r}

plot_marginals <- marginals %>%
  pivot_longer(cols = c("marg_null", "marg_long"), names_to = "Distribution", values_to = "Prob") %>%
  mutate(Distribution = recode(Distribution, marg_null = "null", marg_long = "long")) %>%
  mutate(Distribution = factor(Distribution, levels = c("null", "long")))

ggplot(plot_marginals, aes(x = mu_val, y = Prob, color = Distribution, fill=Distribution)) +
  geom_line(linewidth = 0.8) +
  geom_area(alpha = 0.3, position="identity") +
  facet_wrap(~prior, labeller = label_both, ncol = 1) +
  scale_color_manual(values = c("null" = "gray30", "long" = "blue")) +
  scale_fill_manual(values = c("null" = "gray50", "long" = "blue")) +
  theme_bw() +
  labs(title = "marginal P(mu|R)", x = "mu", y = "P(mu|R)") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
```

```{r}
final_means <- posterior %>%
  group_by(mu, Distribution) %>%
  summarise(mean_O = sum(O * Prob))

final_plot <- ggplot(posterior, aes(x = O, y = Prob, fill = Distribution, color = Distribution)) +
  geom_area(alpha = 0.3, position = "identity") + 
  geom_line(linewidth = 0.8) +
  geom_vline(data = final_means, aes(xintercept = mean_O, color = Distribution), 
             linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~mu, labeller = label_both) +
  scale_fill_manual(values = c("null" = "gray50", "long" = "blue")) +
  scale_color_manual(values = c("null" = "gray30", "long" = "blue")) +
  theme_bw() +
  labs(title = "Posteriors for null vs long", x = "Opinion (O)", y = "P(O|R)")
print(final_plot) 
```
What I'd like to see (what I had in the original graphs, and what showed up in the experiment) is that the grey curve (null) should be similar to 
L1's own prior. So for hyperprior 0.2, I'd expect something with a mean around 0.2. But that's clearly not what we're getting here... and I wonder if
it is a reversal of the backfire effect, where low prior with no explanation triggers an inference that the speaker expected a HIGH prior-- in 
the model at least. my instict is that would not be consistent with the data, but i need to think about it and i'm fading a little


