---
title: "rsa_april"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
V_vals <- seq(0.0001, 0.9999, length.out = 2000) 

#grid of beta params
mu_vals <- seq(0.01, 0.99, length.out = 20)
nu_vals <- seq(1, 10, length.out = 10)

sample_mu_vals <- c(0.25, 0.5, 0.75)


theta_vals <- seq(0.05, 0.95, length.out = 46)


cost <- c(null = 0, long = 0.5) 
alpha <- 2
```

Literal listener:

$$
P_{L_0^{\mu, \nu}}(V|R) \propto P(R|V)\cdot P(V) \\
=  P(R|V)\cdot Beta(V; \mu \nu, (1-\mu)\nu)
$$

```{r}
prior_L0 <- function(mu, nu) {
  a <- mu*nu
  b <- (1-mu) * nu
  dbeta(V_vals, a, b)
}

likelihood_L0 <- function(R) {
  if(R == "null") rep(1, length(V_vals)) else V_vals^2
}

lit_listener <- function(mu, nu, R) {
  prior <- prior_L0(mu, nu)
  prior <- prior / sum(prior)
  likelihood <- likelihood_L0(R)
  unnorm_posterior <- prior * likelihood
  posterior <- unnorm_posterior / sum(unnorm_posterior)
  data.frame(V = V_vals, 
             Prior = prior, 
             Posterior = posterior,
             UnnormPosterior = unnorm_posterior,
             mu = mu, 
             nu = nu, 
             R = R)
}
```

```{r}

L0_data <- lapply(mu_vals, function(m) 
            lapply(nu_vals, function(n) lit_listener(m, n, "long"))) %>%
            bind_rows() %>%
            pivot_longer(cols = c("Prior", "Posterior"), names_to = "Distribution",values_to ="Prob")

figure1_data <- L0_data  %>%
  filter(Distribution == "Posterior") %>%
  group_by(mu, nu) %>%
  summarise(post_mean = sum(V * Prob), .groups = "drop") %>%
  mutate(delta_means = post_mean - mu)

```

```{r}

figure1_data %>% ggplot(aes(x = mu, y = nu, fill = delta_means)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(title = "Change in L0 expected opinion (V) after long explanation (4R)", x = "Prior mu", y = "Prior nu", fill = "Delta") +
  theme_minimal()

```

The general trend we observe here is that there's always a positive change from the prior to the posterior, and that tends to be magnified when the prior is less favorable on average (low mu) or less confident (low nu). Since the null utterance reproduces the prior, and the long utterance conditions on the evidence, the long utterance always does better than the null utterance for L0.

For good measure, for fixed nu:

```{r}

figure2_data <- lapply(sample_mu_vals, function(m) lit_listener(m, 6, "long")) %>%
            bind_rows() %>%
            pivot_longer(cols = c("Prior", "Posterior"), names_to = "Distribution",values_to ="Prob")

posterior_means <- figure2_data %>%
  filter(Distribution == "Posterior") %>%
  group_by(mu) %>%
  summarise(post_mean = sum(V * Prob))

figure2_plot <- ggplot(figure2_data, aes(x = V, y = Prob, fill = Distribution, color = Distribution)) +
  geom_area(alpha = 0.3, position = "identity") + 
  geom_line(size = 0.8) +
  geom_vline(aes(xintercept = mu), linetype = "dashed", color = "gray30", linewidth = 0.5)+ 
  geom_vline(data = posterior_means, aes(xintercept = post_mean), linetype = "dashed", color = "blue", linewidth = 0.5)+
  facet_wrap(~mu, labeller = label_both, ncol=1) +
  scale_fill_manual(values = c("Prior" = "gray50", "Posterior" = "blue")) +
  scale_color_manual(values = c("Prior" = "gray30", "Posterior" = "blue")) +
  theme_bw() +
  labs(title = "Bayesian update with no pragmatic reasoning", x = "Value", y = "P(V|long-4R; mu, nu)")

print(figure2_plot)
```

Here also we can see that the distribution always shifts to the right for the long explanation compared to the prior (no explanation), but the shift is less when the prior is already high.

```{r}

#cached the lookups, otherwise takes forever to run 
L0_array_null <- array(sapply(nu_vals, function(n)
                        sapply(mu_vals, function(m)
                        lit_listener(m, n, "null")$Posterior)),
  dim = c(length(V_vals), length(mu_vals), length(nu_vals)),
  dimnames = list(V = V_vals, mu = mu_vals, nu = nu_vals)
)

L0_array_long <- array(sapply(nu_vals, function(n)
                        sapply(mu_vals, function(m)
                        lit_listener(m, n, "long")$Posterior)),
  dim = c(length(V_vals), length(mu_vals), length(nu_vals)),
  dimnames = list(V = V_vals, mu = mu_vals, nu = nu_vals)
)


#not needed for speaker but needed for L1 update maybe?
L0_unnorm_null <- array(sapply(nu_vals, function(n)
                        sapply(mu_vals, function(m)
                        lit_listener(m, n, "null")$UnnormPosterior)),
  dim = c(length(V_vals), length(mu_vals), length(nu_vals)),
  dimnames = list(V = V_vals, mu = mu_vals, nu = nu_vals)
)

L0_unnorm_long <- array(sapply(nu_vals, function(n)
                        sapply(mu_vals, function(m)
                        lit_listener(m, n, "long")$UnnormPosterior)),
  dim = c(length(V_vals), length(mu_vals), length(nu_vals)),
  dimnames = list(V = V_vals, mu = mu_vals, nu = nu_vals)
)

mu_nu_theta_grid <- expand.grid(
  theta_i = 1:length(theta_vals), 
  mu_j = 1:length(mu_vals), 
  nu_k = 1:length(nu_vals)
)

mu_nu_theta_grid$theta <- theta_vals[mu_nu_theta_grid$theta_i]
mu_nu_theta_grid$mu <- mu_vals[mu_nu_theta_grid$mu_j]
mu_nu_theta_grid$nu <- nu_vals[mu_nu_theta_grid$nu_k]
```

Pragmatic speaker

$$
P_{L_0^{\mu,\nu}}(V > \theta| R) = \int_\theta^1 P_{L_0^{\mu, \nu}}(V|R)  dV
$$

$$
\mathbb{U}_{S_1}(R; \theta, \mu, \nu) = \underbrace{\log P_{L_0^{\mu, \nu}}(O > \theta|R)}_{\textbf{-surprisal}} - \text{cost(R)}
$$

$$
P_{S_1}(R|\theta, \mu, \nu) = \text{softmax}(\alpha \cdot \mathbb{U}_{S_1}(R; \theta, \mu))
$$

```{r}

s_probs <- mapply(function(i, j, k) {
  
  #integral
  p_null <- sum(L0_array_null[V_vals > theta_vals[i], j, k])
  p_long <- sum(L0_array_long[V_vals > theta_vals[i], j, k])
  
  #informativity 
  inf_null <- log(p_null + 1e-10)
  inf_long <- log(p_long + 1e-10)
  
  #utility
  util_null <- inf_null - cost["null"]
  util_long <- inf_long - cost["long"]

  #softmax
  null_numer <- exp(alpha * util_null)
  long_numer <- exp(alpha * util_long)
  denom <- null_numer + long_numer
  
  return(c(null_numer/denom, long_numer/denom, p_null, p_long, inf_null, inf_long))
  
}, mu_nu_theta_grid$theta_i, mu_nu_theta_grid$mu_j, mu_nu_theta_grid$nu_k)

mu_nu_theta_grid$prob_null <- s_probs[1, ]
mu_nu_theta_grid$prob_long <- s_probs[2, ]
mu_nu_theta_grid$int_null <- s_probs[3, ]
mu_nu_theta_grid$int_long <- s_probs[4, ]
mu_nu_theta_grid$inf_null <- s_probs[5, ]
mu_nu_theta_grid$inf_long <- s_probs[6, ]



```

```{r}
speaker_plot <- mu_nu_theta_grid %>%
  mutate(int_diff = int_long - int_null) %>%
  mutate(inf_diff = inf_long - inf_null) %>%
  mutate(prob_diff = prob_long - prob_null) %>%
  filter(theta_i == 10)

  ggplot(speaker_plot, aes(x = mu, y = nu, fill = prob_diff)) +
    geom_tile() +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, 
      breaks = c(min(mu_nu_theta_grid$prob_diff), 0, max(mu_nu_theta_grid$prob_diff)),
      labels = c("Prefers null (0R)", "", "Prefers long (4R)")) +
    labs(title = "Pragmatic speaker utterance preference", subtitle = "P(long|theta, mu, nu) - P(null| theta, mu, nu)",  x = "mu", y = "nu", fill = "") +
    theme_minimal()

```

These results are as expected. The speaker prefers the long explanation for lower values of prior mu, and the null explanation for higher value of prior mu.

Further, the long explanation survives for higher values of prior mu when prior nu is lower (weakly opinionated prior).

```{r}

prior_theta <- dnorm(theta_vals, mean = 0.5, sd = 0.2) 
prior_theta <- prior_theta / sum(prior_theta) 

mean_mu = 0.2
mean_nu = 2
k = 1
  
hyperprior_mu <- dbeta(mu_vals, mean_mu * 3, (1 - mean_mu) * 3)
hyperprior_mu <- hyperprior_mu / sum(hyperprior_mu)

hyperprior_nu <- dgamma(nu_vals, shape = mean_nu * k, rate = k) 
hyperprior_nu <- hyperprior_nu / sum(hyperprior_nu)

#priors p(mu), p(nu), p(theta) for each mu,nu,theta combination 
p_mu <- hyperprior_mu[mu_nu_theta_grid$mu_j]
p_nu <- hyperprior_nu[mu_nu_theta_grid$nu_k]
p_theta <- prior_theta[mu_nu_theta_grid$theta_i]

#joint probability of each mu,nu,theta combination
joint_null_unnorm <- mu_nu_theta_grid$prob_null * p_mu * p_nu * p_theta
joint_long_unnorm <- mu_nu_theta_grid$prob_long * p_mu * p_nu * p_theta

mu_nu_theta_grid$joint_null <- joint_null_unnorm / sum(joint_null_unnorm)
mu_nu_theta_grid$joint_long <- joint_long_unnorm / sum(joint_long_unnorm)

#marginalize out theta to get L1 probs of possible (mu,nu) pairs given utterance
marginals <- mu_nu_theta_grid %>%
  group_by(mu_j, nu_k) %>%
  summarize(marg_null = sum(joint_null), marg_long = sum(joint_long), .groups = "drop")

mu_nu_theta_grid <- mu_nu_theta_grid %>%
  group_by(mu_j, nu_k) %>%
  mutate(marg_null = sum(joint_null), marg_long = sum(joint_long)) %>%
  ungroup()

```

```{r}
ggplot(mu_nu_theta_grid, aes(x = mu, y = nu, fill = marg_null)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    labs(title = "L1 infers prior (mu, nu) from null explanation", x = "mu", y = "nu", fill = "") +
    theme_minimal()

ggplot(mu_nu_theta_grid, aes(x = mu, y = nu, fill = marg_long)) +
    geom_tile() +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
    labs(title = "L1 infers prior (mu, nu) from long explanation",  x = "mu", y = "nu", fill = "") +
    theme_minimal()
  
```

Some of this makes sense– when the pragmatic listener hears the long explanation, he tends to guess lower values of mu, and when he hears the null explanation, he tends to guess higher values of mu (this is the RQ1 effect– participants say the speaker expected the listener to have an less favorable rating in the long condition vs the null). We also see some influence of the prior here– he tends not to guess the more extreme values of mu/nu.

However, I don't think the effect is moderated by nu based on plotting? which doesn't fit my predictions and i dont understand how the final plot is basically working without that?

```{r}

#see how different combinations of the hyperpriors affect the inference
hyperparams <- expand.grid(mean_mu = mu_vals, mean_nu = nu_vals)

#lot of copied code i'll edit it down later
L1_probs <- mapply(function (mean_mu, mean_nu) {
  
  hyperprior_mu <- dbeta(mu_vals, mean_mu * 5, (1 - mean_mu) * 5)
  hyperprior_mu <- hyperprior_mu / sum(hyperprior_mu)
  
  hyperprior_nu <- dgamma(nu_vals, shape = mean_nu * k, rate = k) 
  hyperprior_nu <- hyperprior_nu / sum(hyperprior_nu)
  
  #priors p(mu), p(nu), p(theta) for each mu,nu,theta combination 
  p_mu <- hyperprior_mu[mu_nu_theta_grid$mu_j]
  p_nu <- hyperprior_nu[mu_nu_theta_grid$nu_k]
  p_theta <- prior_theta[mu_nu_theta_grid$theta_i]
  
  #joint probability of each mu,nu,theta combination
  joint_null_unnorm <- mu_nu_theta_grid$prob_null * p_mu * p_nu * p_theta
  joint_long_unnorm <- mu_nu_theta_grid$prob_long * p_mu * p_nu * p_theta
  
  mu_nu_theta_grid$joint_null <- joint_null_unnorm / sum(joint_null_unnorm)
  mu_nu_theta_grid$joint_long <- joint_long_unnorm / sum(joint_long_unnorm)
  
  #marginalize out theta to get L1 probs of possible mu nu pairs given utterance
  marginals <- mu_nu_theta_grid %>%
    group_by(mu_j, nu_k) %>%
    summarize(marg_null = sum(joint_null), marg_long = sum(joint_long), .groups = "drop")
  
  marginals <- marginals %>% arrange(nu_k, mu_j)
  
  #get the distributions for V for each utterance 
  #weighing L0 standard bayes theorem output assuming a given pair of mu, nu
  #by the marginal of that mu, nu pair
  L0_null_mat <- matrix(L0_array_null, nrow = length(V_vals)) 
  L0_long_mat <- matrix(L0_array_long, nrow = length(V_vals))
  p_V_null <- L0_null_mat %*% marginals$marg_null
  p_V_long <- L0_long_mat %*% marginals$marg_long

  #get the expected value of each 
  c(EV_null = sum(V_vals * p_V_null),
    EV_long = sum(V_vals * p_V_long))
  
}, hyperparams$mean_mu, hyperparams$mean_nu)

EV_df <- data.frame(
  mu = hyperparams$mean_mu,
  nu = hyperparams$mean_nu,
  EV_null = L1_probs["EV_null", ],
  EV_long = L1_probs["EV_long", ]
)

```

```{r}

ev_plot <- EV_df %>%
  mutate(EV_diff = EV_long - EV_null)

ggplot(ev_plot, aes(x = mu, y = nu, fill = EV_diff)) +
    geom_tile() +
    scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
    labs(title = "L1 expected value diff",  x = "mu", y = "nu", fill = "") +
    theme_minimal()

```

```{r}

ev_plot %>%
  pivot_longer(c(EV_null, EV_long), names_to = "utterance", values_to = "EV") %>%
  mutate(utterance = recode(utterance, EV_null = "Null", EV_long = "Long")) %>%
  ggplot(aes(x = mu, y = nu, fill = EV)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0.5) +
  facet_wrap(~ utterance) +
  labs(title = "L1 expected value by utterance", x = "mu", y = "nu", fill = "") +
  theme_minimal()

```
